---
title: "Extending aion"
author: "N. Frerebeau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: yes
    fig_caption: yes
    toc: true
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Extending aion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(aion)
```

**aion** offers a simple API that can be extended and used by other specialized packages.

# Calendars and Dates

The following example is used to build a simple solar calendar with 365 days each year and no leap-year rule. This is the ancient Egyptian calendar. You will find full details of the calculations and detailed explanations in @reingold2018 (p. 29).

You can define additional calendars by creating S4 classes that inherit from the `TimeScale` class exported by **aion**:

```{r}
## Egyptian calendar
E <- setClass(
  Class = "EgyptianCalendar",
  prototype = list(
    name = "Egyptian",
    fixed = -272787,
    direction = 1L,
    year = 365
  ),
  contains = "TimeScale"
)
```

Once the calendar has been defined, you need to build methods for converting *rata die* to and from this calendar:

```{r}
## Convert Egyptian dates to rata die
## NB: this method MUST return a RataDie object
setMethod(
  f = "fixed",
  signature = c(
    year = "numeric",
    month = "numeric",
    day = "numeric", 
    calendar = "EgyptianCalendar"
  ),
  definition = function(year, month, day, calendar) {
    rd <- calendar_fixed(calendar) + 
      365 * (year - 1) + 
      30 * (month - 1) + 
      day - 1
    
    as_fixed(rd)
  }
)

## Convert rata die to Egyptian dates
## NB: this method MUST return a data.frame
setMethod(
  f = "as_date",
  signature = c(object = "numeric", calendar = "EgyptianCalendar"),
  definition = function(object, calendar) {
    day <- object - calendar_fixed(calendar)
    year <- day %/% 365 + 1
    month <- (day %% 365) %/% 30 + 1
    day <- day - 365 * (year - 1) - 30 * (month - 1) + 1
    
    data.frame(year = year, month = month, day = day)
  }
)

## Convert rata die to Egyptian years
setMethod(
  f = "as_year",
  signature = c(object = "numeric", calendar = "EgyptianCalendar"),
  definition = function(object, calendar, ...) {
    (object - calendar_fixed(calendar)) %/% 365 + 1
  }
)
```

Now you can use your calendar:

```{r}
## Create a calendar object
cal <- E()

## Convert 161/7/15 in rata die
fixed(
  year = 161,
  month = 7,
  day = 15, 
  calendar = cal
)

## Convert -214193 r.d. to an Egyptian date
as_date(-214193, calendar = cal)
```

The definition of new calendars, combined with the Julian and Gregorian calendars already included in **aion**, allows you to build conversion tools:

```{r}
## Build a conversion function from Gregorian CE years to Egyptian years
Gregorian_to_Egyptian <- convert(CE(), E())

## Convert 2023 (Gregorian) to the Egyptian calendar
Gregorian_to_Egyptian(2023)
```

# Time Series

A time series object is simply an $n \times m \times p$ `array`, with $n$ being the number of observations, $m$ being the number of series and with the $p$ columns of the third dimension containing extra variables for each series. This `array` comes with an extra `time` slot that store the observations times expressed in *rata die*. You can create classes that inherits from the `TimeSeries` class.

As an example, you can create a class that represent the results of the calibration of radiocarbon dates (this code comes from the [**ananke**](https://packages.tesselle.org/ananke/) package):

```{r}
.CalibratedAges <- setClass(
  Class = "CalibratedAges",
  slots = c(
    ages = "numeric",     # Stores the radiocarbon ages to be calibrated
    errors = "numeric",   # Store the standard deviation of the radiocarbon ages
    curves = "character"  # Store the name of the calibration curve
  ),
  contains = "TimeSeries"
)
```

All methods defined in **aion** can then be used on objects belonging to this new class (e.g. `plot()`).

# References
